---
title: "Adaptation and resilience"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## PMD (Coderre administration): 

While the PMD is concerned with mitigation of climate change through community emissions reduction and changes to lifestyle, governance and business practices, the PACCAM addresses adaptation exclusively. PACCAM describes what the city administration and the borough governments can do in the short to medium-term to adapt to the coming challenges of climate change, such as more frequent storms, droughts, extreme heat, heavy rain and flooding. The policies for the city administration are presented as commitments, while the borough policies are implemented on a voluntary basis. There is no enforcement mechanism to ensure that these policies are carried out. Progress was monitored in a 2017 plan update and a 2019 report on the plan’s progress. 

The objectives of the PACCAM often overlap with those of the PMD. These include an increase of the tree canopy to 25% and increase of protected areas to 10% of the agglomeration. The plan also includes less specific goals such as the preservation of riverbanks and natural floodplains, protection of biodiversity, and reduction of heat islands. The plan promises various actions on the part of the municipal administration and proposes policy interventions that the boroughs and municipalities in the urban agglomeration can undertake on a voluntary basis. These policies, whether for the administration or the boroughs, are generally limited in scope, for instance, the city promising to “try out” replacements for mineral abrasives, and suggestions to “make the most” out of green infrastructure. Boroughs are enjoined to increase their slates of winter and summer outdoor activities, regularly cut their grass and regularly inspect buildings and sewer infrastructure. Needless to say at this point, the 246-page list of adaptation measures contains some bloat. While none would object to its measures, the PACCAM focuses on suggesting actions which constitute basic good governance, and essentially delegates the task of climate change adaptation to the boroughs, to be undertaken at their pace, which may be variable and subject to budgetary [...]

## Plan climat (Plante administration): 

Devote 10 to 15% of the city’s ten-year capital budget to adaptation measures. Encourage the community-level development of resilience hubs. Adapt Montreal’s construction and renovation bylaws based on priorities for ecological transition involving, among other things, energy efficiency criteria and architectural criteria for resilience. Improve the resilience and energy efficiency of buildings under the AccèsLogis program. 

## Discussion

The PACCAM and Plan climat have very different understandings of the management of adaptation and resilience. PACCAM favours a voluntary, piecemeal approach whereby boroughs take responsibility for adaptation at their own pace, with oversight and monitoring from the municipal administration. The Plan climat suggests a more direct plan of subsidy, regulation and capital investment. We are likely in the midst of a sea change in terms of climate planning. In the Montreal’s first plans, the concept of adaptation was not only absent, it had not yet been articulated in any type of discourse. By the time of the PMD and PACCAM, adaptation was present as a concept, but urgency was minimal: voluntary measures, carried out gradually by the city administration and the boroughs. By 2020, adaptation (sometimes reworked as resilience) was at the top of the list of priorities. An investment of 10-15% of the city’s capital budget for adaptation would be the single largest climate policy ever implemented in terms of material impact. Possibly influenced by the city’s recent membership in the 100 Resilient Cities Network (100 RC) and its 2018 resiliency plan, the Plan climat 2020-2030 contains more adaptation and resilience strategies than any previous Montreal main climate plan (excluding the PACCAM). This is unsurprising as the department that emerged from Montreal’s involvement in 100 RC, the municipal resilience office, was quickly renamed to the office of ecological transition and resilience and played a significant role in the drafting of the Plan climat; the director of ecological transition and resilience, Sidney Ribaux, a key contributor. Adaptation and resilience are now receiving significant direct funding while being directly applied to the built environment and community, instead of being implemented as a patchwork by the boroughs.

## Vulnerability to climate hazards in Montreal

```{r}
renderMapdeck({
      mapdeck(
        style = map_style, token = token_climate_risk,
        zoom = map_zoom, location = map_location) %>%
        add_polygon(data = {grid %>% 
            dplyr::select(ID, name, name_2, population, 
                          left_var_full = destructive_storms_ind,
                          left_var = destructive_storms_ind_q3) %>% 
            mutate(group = paste(left_var, "- 1")) %>% 
            left_join(colour_borough, by = "group")}, stroke_width = 0, 
          stroke_colour = "#FFFFFF", fill_colour = "fill", update_view = FALSE,
          id = "ID", auto_highlight = TRUE, highlight_colour = "#FFFFFF90")
      })

    # Zoom level
    observeEvent(input$map_view_change$zoom, {
      rv_climate_risk$zoom <- case_when(
        input$map_view_change$zoom >= 12 ~ "DA",
        input$map_view_change$zoom >= 10.5 ~ "CT",
        TRUE ~ "borough")
      })
    
    # Left variable server
    var_left_climate_risk <- select_var_server("left", var_list_climate_risk)
    
    # String to fetch maps and data
    df <- reactive({if (input$grid) "grid" else rv_climate_risk$zoom})
    
    # Right variable server
    var_right_climate_risk <- compare_server("climate_risk", var_list_canale, 
                                             df)

    # Data
    data_climate_risk <- data_server("climate_risk", var_left_climate_risk, 
                                     var_right_climate_risk, 
                                     df, reactive(rv_climate_risk$zoom))
```


